---
title: "HW1"
format: html
editor: visual
---

```{r}
# Mapping rates data
samples <- c("SRR11647686", "SRR11647689", "SRR11647692", 
             "SRR11647696", "SRR11647699", "SRR11647702")

mapping_rates <- c(96.67, 96.50, 96.91, 96.83, 96.86, 96.00)

# Create dataframe
df <- data.frame(Sample = samples, MappingRate = mapping_rates)

# Plot
library(ggplot2)

ggplot(df, aes(x = Sample, y = MappingRate)) +
  geom_bar(stat = "identity", fill = '#001C57') +
  geom_text(aes(label = paste0(MappingRate, "%")), 
            vjust = -0.5, size = 3) +
  coord_cartesian(ylim = c(90, 100)) +
  ylim(0, 100) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Mapping Rate per Sample",
       x = "Sample",
       y = "Mapping Rate (%)") +
  guides(fill = "none")

ggsave("mapping_rate_plot.png", width = 6, height = 4)
```

Clipping profile

```{r}

setwd("C:/Users/avgus/Desktop/Naujas aplankas/clipping_profile")

library(gridExtra)
library(ggplot2)

pdf <- function(...) invisible(NULL)
dev.off <- function(...) invisible(NULL)

samples <- c("SRR11647686", "SRR11647689", "SRR11647692", 
             "SRR11647696", "SRR11647699", "SRR11647702")

all_plots <- list()

for(s in samples){
  source(paste0(s, ".clipping_profile.r"))
  
  # R1
  r1_nonclip_pct <- r1_nonclip_count*100/(r1_clip_count + r1_nonclip_count)
  df_r1 <- data.frame(position = read_pos, nonclip_pct = r1_nonclip_pct)
  all_plots[[paste0(s, "_R1")]] <- ggplot(df_r1, aes(x = position, y = nonclip_pct)) +
    geom_line(color = "blue") +
    geom_point(color = "blue", size = 1) +
    coord_cartesian(ylim = c(94, 100)) +
    theme_classic() +
    theme(plot.title = element_text(face = "bold")) +
    labs(title = paste0(s, " R1"), x = "Position (read-1)", y = "Non-clipped %")
  
  # R2
  r2_nonclip_pct <- r2_nonclip_count*100/(r2_clip_count + r2_nonclip_count)
  df_r2 <- data.frame(position = read_pos, nonclip_pct = r2_nonclip_pct)
  all_plots[[paste0(s, "_R2")]] <- ggplot(df_r2, aes(x = position, y = nonclip_pct)) +
    geom_line(color = "blue") +
    geom_point(color = "blue", size = 1) +
    coord_cartesian(ylim = c(94, 100)) +
    theme_classic() +
    theme(plot.title = element_text(face = "bold")) +
    labs(title = paste0(s, " R2"), x = "Position (read-2)", y = "Non-clipped %")
}

combined <- arrangeGrob(grobs = all_plots, nrow = 4, ncol = 3)
ggsave("clipping_profile_all.png", combined, width = 15, height = 16, dpi = 300)


```

inner distance

**Inner distance** = the gap between the two paired reads on the genome. Imagine your paired-end reads like this:

Read1 ----\> \<---- Read2

\[gap\]

That gap in the middle is the inner distance — it represents the size of the RNA fragment that was sequenced.

What the histogram shows:

X axis — inner distance in base pairs (bp) Y axis — how many read pairs have that inner distance Red line — the distribution curve

Think of it like this:

Your RNA fragment gets sequenced from **both ends**:

```         
Original RNA fragment: |---------------------------------------|  Read 1 sequenced from left: |======>  Read 2 sequenced from right:                             <======|  The gap in the middle = inner distance:         |_________________|
```

**So inner distance is simply:**

> *How much of the original RNA fragment is LEFT UNSEQUENCED in the middle*

**Examples:**

-   **Positive inner distance** = gap between reads, middle part not sequenced

-   **Zero** = reads meet exactly in the middle

-   **Negative inner distance** = reads overlap each other

```{r}

setwd("C:/Users/avgus/Desktop/Naujas aplankas/inner_distance")

library(gridExtra)
library(ggplot2)

pdf <- function(...) invisible(NULL)
dev.off <- function(...) invisible(NULL)
write <- function(...) invisible(NULL)

samples <- c("SRR11647686", "SRR11647689", "SRR11647692", 
             "SRR11647696", "SRR11647699", "SRR11647702")

all_plots <- list()

for(s in samples){
  source(paste0(s, ".inner_distance_plot.r"))
  
  frag_mean <- round(mean(fragsize), 1)
  frag_sd <- round(sd(fragsize), 1)
  
  df <- data.frame(fragsize = fragsize)
  
  all_plots[[s]] <- ggplot(df, aes(x = fragsize)) +
    geom_histogram(aes(y = after_stat(density)), 
                   bins = 100, fill = "white", color = "blue") +
    geom_density(bw = 10, color = "red") +
    theme_classic() +
    theme(plot.title = element_text(face = "bold", size = 9)) +
    labs(title = paste0(s, "\nMean=", frag_mean, "; SD=", frag_sd),
         x = "mRNA insert size (bp)",
         y = "Density")
}

combined <- arrangeGrob(grobs = all_plots, nrow = 2, ncol = 3)
ggsave("inner_distance_all.png", combined, width = 15, height = 8, dpi = 300)

```

junctions

**1. `splice_junction.pdf`** — shows junction types:

Known junctions = already annotated in the genome 
Novel junctions = new, not previously annotated 
Partial novel = one side known, one side new

-   High % known junctions = good, reads are mapping to real splice sites
-   High % novel = could be interesting biology OR mapping errors

**2. `splice_events.pdf`** — shows splicing event types:

SE = Skipped exon (most common) RI = Retained intron A3 = Alternative 3' splice site A5 = Alternative 5' splice site MXE = Mutually exclusive exons

```{r}

setwd("C:/Users/avgus/Desktop/Naujas aplankas/junctions")

library(gridExtra)
library(ggplot2)

pdf <- function(...) invisible(NULL)
dev.off <- function(...) invisible(NULL)

samples <- c("SRR11647686", "SRR11647689", "SRR11647692", 
             "SRR11647696", "SRR11647699", "SRR11647702")

events_plots <- list()
junction_plots <- list()

for(s in samples){
  source(paste0(s, ".junction_plot.r"))
  
  # Splice events
  events_df <- data.frame(
    category = c("partial_novel", "complete_novel", "known"),
    value = events
  )
  events_df$label <- paste0(events_df$category, "\n", round(events_df$value, 1), "%")
  
  events_plots[[s]] <- ggplot(events_df, aes(x = "", y = value, fill = category)) +
    geom_bar(stat = "identity", width = 1) +
    coord_polar("y", start = 0) +
    geom_text(aes(label = label), 
              position = position_stack(vjust = 0.5), size = 3) +
    scale_fill_manual(values = c("partial_novel" = "#E74C3C", 
                                  "complete_novel" = "#2ECC71", 
                                  "known" = "#3498DB")) +
    theme_void() +
    theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 9),
          legend.position = "none") +
    labs(title = paste0(s, "\nSplice Events"))
  
  # Splice junctions
  junction_df <- data.frame(
    category = c("partial_novel", "complete_novel", "known"),
    value = junction
  )
  junction_df$label <- paste0(round(junction_df$value, 1), "%")
  
  junction_plots[[s]] <- ggplot(junction_df, aes(x = "", y = value, fill = category)) +
    geom_bar(stat = "identity", width = 1) +
    coord_polar("y", start = 0) +
    geom_text(aes(label = label), 
              position = position_stack(vjust = 0.5), size = 3) +
    scale_fill_manual(values = c("partial_novel" = "#E74C3C", 
                                  "complete_novel" = "#2ECC71", 
                                  "known" = "#3498DB")) +
    theme_void() +
    theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 9),
          legend.position = "none") +
    labs(title = paste0(s, "\nSplice Junctions"))
}

combined_events <- arrangeGrob(grobs = events_plots, nrow = 2, ncol = 3)
ggsave("splice_events_all.png", combined_events, width = 15, height = 8, dpi = 300)

combined_junction <- arrangeGrob(grobs = junction_plots, nrow = 2, ncol = 3)
ggsave("splice_junction_all.png", combined_junction, width = 15, height = 8, dpi = 300)

```

FEATURE COUNTS

```{r}
setwd("C:/Users/avgus/Desktop/Naujas aplankas/feature_counts")

library(ggplot2)

samples <- c("SRR11647686", "SRR11647689", "SRR11647692", 
             "SRR11647696", "SRR11647699", "SRR11647702")

# Data from your summary
assigned        <- c(12963412, 15473607, 15609816, 15480807, 12839027, 13831651)
unmapped        <- c(228574, 274499, 183042, 206135, 149000, 255753)
multimapping    <- c(2268070, 2763639, 2363175, 1881154, 1358358, 2432250)
no_features     <- c(3206581, 2159798, 2614690, 2005150, 2681209, 1722699)
ambiguity       <- c(1057408, 1218044, 1236647, 1090112, 808639, 1350338)

# Create dataframe
df <- data.frame(
  Sample = rep(samples, 5),
  Category = rep(c("Assigned", "Unmapped", "MultiMapping", 
                   "No Features", "Ambiguity"), each = 6),
  Count = c(assigned, unmapped, multimapping, no_features, ambiguity)
)

# Calculate percentages
df$Total <- rep(assigned + unmapped + multimapping + no_features + ambiguity, 5)
df$Percentage <- df$Count / df$Total * 100

# Plot
ggplot(df, aes(x = Sample, y = Percentage, fill = Category)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = ifelse(Percentage > 3, paste0(round(Percentage, 1), "%"), "")),
            position = position_stack(vjust = 0.5),
            size = 3, color = "black", fontface = "bold") +
  scale_fill_manual(values = c(
    "Assigned"     = "#e89eb8",
    "Unmapped"     = "#332932",
    "MultiMapping" = "#bcbd22",
    "No Features"  = "#efb7b5",
    "Ambiguity"    = "#fcf7ed"
  )) +
  theme_classic(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.title = element_text(face = "bold")
  ) +
  labs(title = "featureCounts Assignment Rates",
       x = "Sample",
       y = "Percentage (%)",
       fill = "Category")

ggsave("featurecounts_assignment.png", width = 10, height = 6, dpi = 300)
```
